---
title: "Assignment 2: Coding Basics"
author: "Yilin Zhong"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. We generate a sequence of numbers from one to 100, increasing by fours and assign this sequence a name of hundred_sequence.
hundred_sequence<-seq(1,100,4)
hundred_sequence
#2. We compute the mean and median of this sequence and assign them a name
means<-mean(hundred_sequence)
means
medians<-median(hundred_sequence)
medians
#3. In here we ask R to determine whether the mean is greater than the median
means > medians
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r vector}
name<- c("John", "Vex", "Kat", "Mandy") #character
class(name)

score<- c(45,100,85,79) #numeric
class(score)

pass<- c("FALSE","TRUE","TRUE","TRUE") #character
class(pass)

```

```{r data frame}
name.df<-as.data.frame(name)
score.df<-as.data.frame(score)
pass.df<-as.data.frame(pass)

exam.df<-cbind(name.df,score.df,pass.df)
exam.df <- data.frame("student_name"=name,"exam_score"=score, "passing_status"=pass)

exam.df
```


9.  QUESTION: How is this data frame different from a matrix?

> Answer:Data frame is different from a matrix by matrix can only contain a single class of data, while data frame can contain different classes of data. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
status1<-function(x){
  if(x>=50) {
    x=TRUE
  } else{
    x=FALSE
  }
}

#report1<-status1(score);report1 

status2<-function(x) {
  ifelse(x>=50,TRUE,FALSE)
}
report2<-status2(score);report2


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Only 'ifelse' function works because 'ifelse' function checks each element in a vector one at a time. However, 'if' and 'else' function can only check one element in a vector at one time, but our code try to check every element in the vector at once, so the function can not be performed.
