---
title: "Assignment 3: Data Exploration"
author: "Yilin Zhong"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r workind directory and loading files,  message = FALSE}
library(tidyverse)
library(lubridate)
getwd()
Neonics<-read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",stringsAsFactors = T)
Litter<-read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",stringsAsFactors = T)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:We are interested in the ecotoxicology of neonicotionoids on insects because neonicotionoids is a class of insecticides, which contains toxic chemicals to kill insects. We want to know the toxicity that this class of insecticide contains to determine their effects on the environment when people applied it in the fields. The ecotoxicology of neonicotionoids allow us to determine the amount people should use while have the least toxic effects on the environment and people.  

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:We are interested in studying litter and woody debris that falls to the ground in forests because they play an important role on carbon budget and nutrient cycling. They are a source of energy in the aquatic ecosystems. In addition, they provide habitat for terretrial and aquatic organisms. Moroever, they determines the structure and roughness of their area, which influence water flow and sediment transports. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: Litter and woody debris sampled as part of the NEON network by elevated and ground traps, respectively.
 1.Litter and fine woody debris sampling took place at terrestrial NEON sites that contain woody vegetation >2m tall. Together with most of NEON’s plant productivity measurements, sampling for this product occurs only in tower plots. Locations of tower plots are choosen randomly within the 90% flux footprint of the primary and secondary airsheds.
 2.One litter trap pair, which includes one elevated trap and one ground trap, is deployed for every 400 m^2 plot area, led to 1-4 trap pairs per plot.
 3. Plot edges must be separated by a distance 150% of one edge of the plot; plot centers must be greater than 50 meter from large paved roads and plot edges must be 10m from two-track dirt roads; plot centers must be 50 meter from buildings and other non-NEON infrastructure; streams larger than 1 meter must not intersect plots.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r dimension}
dim(Neonics) #dimension is 4623 observations and 30 variables
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r summary on effects}
summary(Neonics$Effect) #counts the total numbers in each effect category
```

> Answer:The most common effects that are studied are mortality and population, which each exceeds 1000. These effects are speicifically be of interest because mortality reveals the measurements and endpoints where the cause of death is by direct action of the chemical. We want to know how many deaths are cause by the chemical and using these info to decide further action if needed. As for population, it reveals the measurements and endpoints related to a group of organisms or plants of the same species occupying the same area at a given time. We interested in population because we want to know what groups of organisms or plants are in the specific location at a sepcific time to see if there are exposure to chemicals. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r summary of commonly studied species}
sort(summary(Neonics$Species.Common.Name)) #counts the total numbers in each studied species

```

> Answer:The six most commonly studied species in the dataset are Honey bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honeybee. These species are all certain types of bees. They might be of interest over other insects because they have a role of pollinators that carry large grains of pollen between plants. Since they carry pollens, they also carry the insecticides that we are interested in our study, which they can expose the chemicals to other species as they fly around. This can work the other way as well, as they fly around, they can be easily exposed to the chemicals as they often have close contacts with plants. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r class for conc author}
class(Neonics$Conc.1..Author.) #factor class
```

> Answer:The class for this column in the dataset is factor. It is not numeric because not all the values are in numbers. For instance, it contains NR, which means the number is not reported. Numeric class is not the best class to use in this case. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r frequency plot by year}
#plot of the number of studies conducted by publication year
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50) 
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r test location}
#plot of the number of studies conducted by publication year with color aesthetic that shows 
#different Test.Location and displayed as different colors.
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color= Test.Location), bins = 50) 
  
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common locations are lab and field nature. These two locations remain the most common locations at all time. However, in some years, field nature exceeds lab as the most common location, such as from year 1997 to 2000, and 2008 to 2010. For the rest of the timeframe, lab is the most common location.  

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r bar graph for endpoint}
#bar graph of Endpoint counts
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer:The two most common endpoints are LOEL and NOEL. LOEL is the lowest observable effect level, which means the lowest dose (concentration) producing effects that were significantly different from responses of controls. NOEL is the no observable effect level, which means the highest dose producing effects not significantly different from responses of controls according to the author's reported statisical test.  

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r date format}
class(Litter$collectDate) #the class is factor
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
class(Litter$collectDate) #now is date

unique(Litter$collectDate) # sampled on 2018-08-02 and 2018-08-30

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r plots}
unique(Litter$plotID)
summary(Litter$plotID)
```

> Answer: There are 12 plots sampled at Niwot Ridge. Unique function eliminate duplicate values, which shows how many unique plots are sampled at Niwot Ridge. As for summary, it shows how many collections took place at each plot. 


14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r bar graph functional group}
# bar graph of functionalGroup counts
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r drymass plots}
#boxplot of dryMass by functionalGroup and we rotate the x label for better orientation
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
#violin plot of dryMass by functionalGroup and we rotate the x label for better orientation
ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass), 
              draw_quantiles = c(0.25, 0.5, 0.75))+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:Boxplot is a more effective visualization option than the violin plot in this case because the sample size is relatively small, the density estimations that violin plot usually shows does not work in here. It does not have enough data to show the density estimations.  


What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles have the highest biomass at these sites, and Mixed have the second highest biomass. 
